---
title: "report_charts"
author: "Dominic Jones"
format: html
editor: visual
---

## setup

```{r}
#| echo: false
#setup

#load packages
library(targets)
library(dplyr)
library(tidyr) 
library(ggplot2)
library(patchwork)
library(grattantheme)
library(forcats)

#library(waterfalls)


```

# Households are set to save in the transition: fixing the safeguard will cost them little

## 1.Retail prices will look like this for each of the options:

setup

```{r}
#|echo: false

tar_load(c(jacobs_retail_prices_reference_case,
           jacobs_retail_prices_1_5_opt1,
           jacobs_retail_prices_1_5_opt2)) 

tar_load(household_connections)

source("R/helpers.R")
  
retail_prices <- bind_rows(jacobs_retail_prices_reference_case,
           jacobs_retail_prices_1_5_opt1,
           jacobs_retail_prices_1_5_opt2) %>% 
  mutate(scenario = fct_case_when(scenario == "Ref" ~ "Reference case",
                                  scenario == "1_5_Opt1" ~ "RET 1.5 C",
                                  scenario == "1_5_Opt2" ~ "Safeguard 1.5 C"))

chart_palette <- c("Reference case" = grattan_red,
                   "RET 1.5 C" = grattan_yellow,
                   "Safeguard 1.5 C" = grattan_orange)
```

Residential prices by state and scenario over time:

```{r}
#|echo: false

retail_prices %>% 
  filter(market == "Residential",
         consumption_export == "Consumption",
         state %nin% c("ACT", "NT")) %>% 
  ggplot(aes(x = year, 
             y = (c_kwh / 100) * 1000, #convert to dollars per mwh
             colour = scenario)) +
  geom_line() +
  facet_wrap(~state) +
  grattan_y_continuous(labels = scales::dollar_format()) +
  scale_colour_manual(values = chart_palette) +
  theme_grattan() +
  labs(title = "In most states, prices are higher under a carbon price, but the Safeguard option quickly converges",
       subtitle = "Dollars per MWh ($2025)",
       x = "",
       y = "")
  
```

Electricity bills by state and scenario

```{r}

tar_load(weighted_average_electricity_costs)

net_average_electricity_costs <- weighted_average_electricity_costs %>% 
  mutate(category = "Electricity") %>% 
  group_by(year, state, scenario, electrification, category) %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  ungroup() %>% 
  mutate(scenario = fct_case_when(scenario == "Ref" ~ "Reference case",
                                  scenario == "1_5_Opt1" ~ "RET 1.5 C",
                                  scenario == "1_5_Opt2" ~ "Safeguard 1.5 C"))


net_average_electricity_costs %>% 
  filter(electrification == T) %>% 
  ggplot(aes(x = year, y = average_cost_dollars, colour = scenario)) +
  geom_line() +
  facet_wrap(~state) +
  grattan_y_continuous(limits = c(0,4500), labels = scales::dollar_format()) +
  scale_colour_manual(values = chart_palette) +
  theme_grattan() +
  labs(title = "Household bills are set to rise across scenarios",
       subtitle = "Household electricity bills ($2025)",
       x = "",
       y = "")
  

```

save chart

```{r}
#| echo: false

grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/electricity_bills_by_scenario.pdf",
             object = ggplot2::last_plot())

```

## 2. But electricity is just one part of the household energy bill 

setup

```{r}

rm(list = ls())
source("R/helpers.R")

#load data
tar_load(weighted_average_electricity_costs)
tar_load(average_gas_costs)
tar_load(average_petrol_costs)
tar_load(household_connections)

chart_fill_palette <- c(
  "Petrol" = grattan_red,
  "Gas" = grattan_orange, 
  "Electricity" = grattan_yellow,
  "Electrification savings" = "transparent",
  "Total costs without electrification" = grattan_black
)

chart_colour_palette <- c(
  "Petrol" = grattan_red,
  "Gas" = grattan_orange, 
  "Electricity" = grattan_yellow,
  "Electrification savings" = grattan_black,
  "Total costs without electrification" = grattan_black
)


net_average_electricity_costs <- weighted_average_electricity_costs %>% 
  mutate(category = "Electricity") %>% 
  group_by(year, state, scenario, electrification, category) %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  ungroup()

#cross join petrol and gas costs with our scenarios (values are the same across scenarios)
petrol_and_gas_costs <- bind_rows(average_gas_costs,
                                  average_petrol_costs) %>% 
  cross_join(tibble(scenario = net_average_electricity_costs %>% 
                      select(scenario) %>% 
                      unique() %>% 
                      pull()))

```

Electricity is just part of the energy budget for households:

```{r}
#|echo: false

wallet_chart_data_nem <- bind_rows(net_average_electricity_costs, 
                                   petrol_and_gas_costs) %>% 
  left_join(household_connections, by = join_by(year, state)) %>% 
  group_by(year, electrification, category) %>% 
  filter(year == 2025,
         #wallet is the same across scenarios in 2025
         scenario == "Ref",
         electrification == T,
         #only include nem states
         state != "WA") %>% 
  summarise(average_cost_dollars = weighted.mean(average_cost_dollars, 
                                                 connections)) 

wallet_chart_data_nem %>% 
  ggplot(aes(reorder(x = category, -average_cost_dollars), 
             y = average_cost_dollars, fill = category, colour = category)) +
  geom_col() +
  theme_grattan() +
  scale_colour_manual(values = chart_fill_palette) +
  scale_fill_manual(values = chart_fill_palette) +
  grattan_y_continuous(labels = scales::dollar_format()) +
    labs(title = "Electricity forms a small part of total household energy costs",
       subtitle = 'Average annual household energy costs in NEM states, by fuel source, 2025',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")
```

Save chart;

```{r}
#| echo: false

grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/average_energy_wallet.pdf",
             object = ggplot2::last_plot())
```

Table 1, energy consumption and bill for every state

Average household energy bills are set to decline across all scenarios (?)

```{r}

chart_scenario_palette <- c("Reference case" = grattan_red,
                   "RET 1.5 C" = grattan_yellow,
                   "Safeguard 1.5 C" = grattan_orange)

household_energy_bills_time <- bind_rows(net_average_electricity_costs, 
                                   petrol_and_gas_costs) %>% 
  filter(electrification == T,
         year >= 2025) %>% 
  group_by(year, state, scenario) %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  ungroup() %>% 
    mutate(scenario = fct_case_when(scenario == "Ref" ~ "Reference case",
                                  scenario == "1_5_Opt1" ~ "RET 1.5 C",
                                  scenario == "1_5_Opt2" ~ "Safeguard 1.5 C"))

household_energy_bills_time %>% 
  ggplot(aes(x = year, y = average_cost_dollars, colour = scenario)) +
  geom_line() +
  facet_wrap(~state) +
  grattan_y_continuous(limits = c(0,8500), labels = scales::dollar_format()) +
  scale_colour_manual(values = chart_scenario_palette) +
  theme_grattan() +
  labs(title = "Average household energy bills are set to decline across all scenarios",
       subtitle = "Total household energy bills ($2025)",
       x = "",
       y = "")

```

save chart:

```{r}
#|echo: false
grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/total_energy_bill_time.pdf",
             object = ggplot2::last_plot())

```
