---
title: "report_charts"
author: "Dominic Jones"
format: html
editor: visual
---

## setup

```{r}
#| echo: false
#setup

#load packages
library(targets)
library(dplyr)
library(tidyr) 
library(ggplot2)
library(patchwork)
library(grattantheme)
library(forcats)

#library(waterfalls)


```

# Households are set to save in the transition: fixing the safeguard will cost them little

## 1.Retail prices will look like this for each of the options:

setup

```{r}
#|echo: false

tar_load(c(jacobs_retail_prices_reference_case,
           jacobs_retail_prices_1_5_opt1,
           jacobs_retail_prices_1_5_opt2,
           jacobs_retail_prices_2_opt1,
           jacobs_retail_prices_2_opt2)) 

tar_load(household_connections)

source("R/helpers.R")
  
retail_prices <- bind_rows(jacobs_retail_prices_reference_case,
           jacobs_retail_prices_1_5_opt1,
           jacobs_retail_prices_1_5_opt2,
           jacobs_retail_prices_2_opt1,
           jacobs_retail_prices_2_opt2) %>% 
  mutate(scenario = fct_case_when(scenario == "Ref" ~ "Reference case",
                                  scenario == "1_5_Opt1" ~ "RET 1.5 C",
                                  scenario == "1_5_Opt2" ~ "Safeguard 1.5 C",
                                  scenario == "2_Opt1" ~ "RET 2 C",
                                  scenario == "2_Opt2" ~ "Safeguard 2 C"),
         carbon_budget = str_match(scenario, "\\d.*C")[,1],
         policy_option = str_match(scenario, ".*[\\s\\d|\\$]") %>% 
           str_remove_all("[\\s|\\.|\\d]"))

chart_palette <- c("Reference" = grattan_red,
                   "RET" = grattan_yellow,
                   "Safeguard" = grattan_orange)
```

Average retail prices across the NEM, weighted by residential retail billable volumes (total residential underlying minus PV-self consumption).

```{r}
#|echo: false

tar_load(annual_electricity_consumption_averages)
tar_load(average_consumer_type_weights)
tar_load(rbs_households)


aggregate_consumption <- annual_electricity_consumption_averages %>% 
  filter(consumption_export == "Consumption",
         electrification == T) %>% 
  mutate(consumer_type = paste0(pv, "_", battery)) %>% 
  left_join(average_consumer_type_weights) %>% 
  left_join(household_connections) %>% 
  mutate(aggregate_consumption_mwh = annual_consumption_kwh * prop * connections / 1000) %>% 
  group_by(year, state) %>% 
  summarise(aggregate_consumption_mwh = sum(aggregate_consumption_mwh))

retail_prices_nem_res <- retail_prices %>% 
  filter(market == "Residential")

    #weight nsw and act tariffs by household numbers in 2020 to get aggregate tariff (this aligns with our connection based weighting of consumption)
  
  retail_prices_nem_res_nsw_act <- retail_prices_nem_res %>% 
    filter(state %in% c("NSW", "ACT")) %>% 
    left_join(rbs_households %>% 
                filter(year == 2020) %>% 
                select(- year)) %>% 
    group_by(year, market, scenario, carbon_budget, policy_option, consumption_export) %>% 
    summarise(c_kwh = weighted.mean(c_kwh, occupied_households)) %>% 
    ungroup() %>% 
    mutate(state = "NSW and ACT")
  
  #add back in
  
  retail_prices_nem_res <- retail_prices_nem_res %>%
    bind_rows(retail_prices_nem_res_nsw_act) %>% 
                filter(state %nin% c("NSW", "ACT"))
  
  
weighted_nem_average_retail_prices <- retail_prices_nem_res %>% 
  left_join(aggregate_consumption) %>% 
  filter(state != "WA",
         state != "NT") %>% 
  group_by(year, scenario, carbon_budget, policy_option) %>% 
  summarise(c_kwh = weighted.mean(c_kwh, aggregate_consumption_mwh)) %>% 
  filter(year >= 2025)
# 
#   #make observations for the reference case for both 1.5 C and 2 C (trajectory is the same bcos unconstrained)
# weighted_nem_average_retail_prices <- weighted_nem_average_retail_prices %>% 
#   filter(!is.na(carbon_budget)) %>% 
# 
#   mutate(carbon_budget = if_else(is.na(carbon_budget), "1.5 C", carbon_budget)) %>%
  


weighted_nem_average_retail_prices %>% 
  ggplot(aes(x = year, y = c_kwh, colour = scenario)) +
  geom_line()

```

Check the our estimates for residential operational served approx equal ESOO forecasts:

This is no longer relevant as Vehicle demand has been scaled differently to AEMO to align with average petrol usage SMVU (but you can see that still broadly aligns)

```{r}
#| echo: false

#operational served = net residential demand
#                   = imports - exports

aggregate_consumption <- annual_electricity_consumption_averages %>% 
  filter(electrification == T) %>% 
  mutate(consumer_type = paste0(pv, "_", battery)) %>% 
  left_join(average_consumer_type_weights) %>% 
  left_join(household_connections) %>% 
  mutate(annual_consumption_t_wh = annual_consumption_kwh * prop * connections / 1e9) %>% 
  group_by(year, state) %>% 
  summarise(annual_consumption_t_wh = sum(annual_consumption_t_wh)) %>% 
  mutate(category = "modelled")


#load ESOO res opertational demand

tar_load(c(esoo_2024_operational_file, wem_esoo_2024_operational_file,
           residential_ev_econsumption, household_connections,
           ev_fleet_data))
  
esoo_res_demand_non_ev <-  read_excel(esoo_2024_operational_file) %>% 
    clean_names() %>% 
    filter(scenario %in% c('Actual', 'Central'),
           parent_category == 'Operational (Sent Out)',
           #only include explicitly residential categories. residential EV will be added on top.
           category %in% c('Residential', 'Electrification'),
           
           #only include residential electrification
           sub_category != 'Business',
           region != 'NEM') %>% 
    rename(state = region) %>% 
    mutate(state = convert_states(state),
           state = if_else(state == 'NSW', 'NSW and ACT', state)) %>% 
    group_by(year, state, category) %>% 
    summarise(annual_consumption_t_wh = sum(annual_consumption_t_wh))  
    



#get western australia data
  wem_esoo_res_demand_non_ev <- read_excel(wem_esoo_2024_operational_file) %>% 
  clean_names() %>% 
  filter(scenario %in% c('Actual', 'Expected (Step Change)'),
         parent_category == 'Operational (Sent Out)',
         #only include explicitly residential categories. residential EV will be added on top.
         category %in% c('Residential', "Electrification"),
         #only include residential electrification
         sub_category != 'Business') %>% 
  mutate(state = "WA") %>% 
  group_by(year, state, category) %>% 
  summarise(annual_consumption_t_wh = sum(annual_consumption_t_wh))
  

  #scale EV demand to total fleet of 1.8 vehicles per household as assumed for our average households
  ev_scale_factor <- ev_fleet_data %>% 
    group_by(year, state) %>% 
    summarise(vehicles_count = sum(vehicles_count)) %>% 
    left_join(household_connections) %>% 
    mutate(vehicles_per_household = vehicles_count / connections,
           scale_factor = 1.8 / vehicles_per_household) %>% 
    select(year, state, scale_factor)
  
  residential_ev_econsumption <- residential_ev_econsumption %>% 
    left_join(ev_scale_factor) %>% 
    mutate(annual_consumption_t_wh = annual_consumption_t_wh * scale_factor) %>% 
    select(-scale_factor)
  
  
  total_esoo_res_demand <- bind_rows(esoo_res_demand_non_ev,
                                     wem_esoo_res_demand_non_ev,
                                     residential_ev_econsumption) %>% 
    group_by(year, state) %>% 
    summarise(annual_consumption_t_wh = sum(annual_consumption_t_wh)) %>% 
  mutate(category = "AEMO forecast")
  
  
  #make chart - Pretty much perfect match !!!
  
  bind_rows(aggregate_consumption,
            total_esoo_res_demand) %>% 
    ggplot(aes(x = year, y = annual_consumption_t_wh, colour = category)) +
    facet_wrap(~state) +
    geom_line()

```

Residential prices by state and scenario over time 1.5 C:

```{r}
#|echo: false

retail_prices %>% 
  filter(market == "Residential",
         consumption_export == "Consumption",
         state %nin% c("ACT", "NT")) %>% 
  ggplot(aes(x = year, 
             y = (c_kwh / 100) * 1000, #convert to dollars per mwh
             colour = scenario)) +
  geom_line() +
  facet_wrap(~state) +
  grattan_y_continuous(labels = scales::dollar_format()) +
  scale_colour_manual(values = chart_palette) +
  theme_grattan() +
  labs(title = "In most states, prices are higher under a carbon price, but the Safeguard option quickly converges",
       subtitle = "Dollars per MWh ($2025)",
       x = "",
       y = "")
  
```

Electricity bills by state and scenario

```{r}

tar_load(weighted_average_electricity_costs)

net_average_electricity_costs <- weighted_average_electricity_costs %>% 
  mutate(category = "Electricity") %>% 
  group_by(year, state, scenario, electrification, category) %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  ungroup() %>% 
  mutate(scenario = fct_case_when(scenario == "Ref" ~ "Reference case",
                                  scenario == "1_5_Opt1" ~ "RET 1.5 C",
                                  scenario == "1_5_Opt2" ~ "Safeguard 1.5 C"))


net_average_electricity_costs %>% 
  filter(electrification == T) %>% 
  ggplot(aes(x = year, y = average_cost_dollars, colour = scenario)) +
  geom_line() +
  facet_wrap(~state) +
  grattan_y_continuous(limits = c(0,4500), labels = scales::dollar_format()) +
  scale_colour_manual(values = chart_palette) +
  theme_grattan() +
  labs(title = "Household bills are set to rise across scenarios",
       subtitle = "Household electricity bills ($2025)",
       x = "",
       y = "")
  

```

save chart

```{r}
#| echo: false

grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/electricity_bills_by_scenario.pdf",
             object = ggplot2::last_plot())

```

## 2. But electricity is just one part of the household energy bill 

setup

```{r}

rm(list = ls())
source("R/helpers.R")

#load data
tar_load(weighted_average_electricity_costs)
tar_load(average_gas_costs)
tar_load(average_petrol_costs)
tar_load(household_connections)

chart_fill_palette <- c(
  "Petrol" = grattan_red,
  "Gas" = grattan_orange, 
  "Electricity" = grattan_yellow,
  "Electrification savings" = "transparent",
  "Total costs without electrification" = grattan_black
)

chart_colour_palette <- c(
  "Petrol" = grattan_red,
  "Gas" = grattan_orange, 
  "Electricity" = grattan_yellow,
  "Electrification savings" = grattan_black,
  "Total costs without electrification" = grattan_black
)


net_average_electricity_costs <- weighted_average_electricity_costs %>% 
  mutate(category = "Electricity") %>% 
  group_by(year, state, scenario, electrification, category) %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  ungroup()

#cross join petrol and gas costs with our scenarios (values are the same across scenarios)
petrol_and_gas_costs <- bind_rows(average_gas_costs,
                                  average_petrol_costs) %>% 
  cross_join(tibble(scenario = net_average_electricity_costs %>% 
                      select(scenario) %>% 
                      unique() %>% 
                      pull()))

```

Electricity is just part of the energy budget for households:

```{r}
#|echo: false

wallet_chart_data_nem <- bind_rows(net_average_electricity_costs, 
                                   petrol_and_gas_costs) %>% 
  left_join(household_connections, by = join_by(year, state)) %>% 
  group_by(year, electrification, category) %>% 
  filter(year == 2025,
         #wallet is the same across scenarios in 2025
         scenario == "Ref",
         electrification == T,
         #only include nem states
         state != "WA") %>% 
  summarise(average_cost_dollars = weighted.mean(average_cost_dollars, 
                                                 connections)) 

wallet_chart_data_nem %>% 
  ggplot(aes(reorder(x = category, -average_cost_dollars), 
             y = average_cost_dollars, fill = category, colour = category)) +
  geom_col() +
  theme_grattan() +
  scale_colour_manual(values = chart_fill_palette) +
  scale_fill_manual(values = chart_fill_palette) +
  grattan_y_continuous(labels = scales::dollar_format()) +
    labs(title = "Electricity forms a small part of total household energy costs",
       subtitle = 'Average annual household energy costs in NEM states, by fuel source, 2025',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")
```

Save chart;

```{r}
#| echo: false

grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/average_energy_wallet.pdf",
             object = ggplot2::last_plot())
```

Table 1, energy consumption and bill for every state

Average household energy bills are set to decline across all scenarios (?)

```{r}

chart_scenario_palette <- c("Reference case" = grattan_red,
                   "RET 1.5 C" = grattan_yellow,
                   "Safeguard 1.5 C" = grattan_orange)

household_energy_bills_time <- bind_rows(net_average_electricity_costs, 
                                   petrol_and_gas_costs) %>% 
  filter(electrification == T,
         year >= 2025) %>% 
  group_by(year, state, scenario) %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  ungroup() %>% 
    mutate(scenario = fct_case_when(scenario == "Ref" ~ "Reference case",
                                  scenario == "1_5_Opt1" ~ "RET 1.5 C",
                                  scenario == "1_5_Opt2" ~ "Safeguard 1.5 C"))

household_energy_bills_time %>% 
  ggplot(aes(x = year, y = average_cost_dollars, colour = scenario)) +
  geom_line() +
  facet_wrap(~state) +
  grattan_y_continuous(limits = c(0,8500), labels = scales::dollar_format()) +
  scale_colour_manual(values = chart_scenario_palette) +
  theme_grattan() +
  labs(title = "Average household energy bills are set to decline across all scenarios",
       subtitle = "Total household energy bills ($2025)",
       x = "",
       y = "")

```

save chart:

```{r}
#|echo: false
grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/total_energy_bill_time.pdf",
             object = ggplot2::last_plot())

```
