---
title: "creating_carbon_budgets"
format: html
editor: visual
execute:
freeze: auto  # re-render only when source changes
---

Load packages etc.

```{r}
#| echo: false

library(dplyr)
library(tidyr) 
library(readr)
library(janitor) 
library(ggplot2)
library(lubridate)
library(readxl)
library(stringr)
library(grattantheme)
library(ggdirectlabel)
library(forcats)
```

Load demand forecast:

```{r}
#| echo: false

#east coast electricity demand
nem_esoo_2024_operational_file <- 'Data/2024 ESOO/2024 ESOO operational (sent out).xlsx'

#west coast electricity demand
# wem_esoo_2024_operational_file <- 'Data/2024 ESOO/2024 WEM ESOO operational (sent out).xlsx'


esoo_demand <- read_excel(nem_esoo_2024_operational_file) %>% 
    clean_names() %>% 
    #here, we select the observed and actual  
    filter(scenario %in% c('Actual', 'Central'),
           parent_category == 'Operational (Sent Out)',
           category != "Operational (Sent Out)",
           region == 'NEM') %>% 
  group_by(year) %>% 
  summarise(annual_consumption_t_wh = sum(annual_consumption_t_wh)) %>% 
  filter(year >= 2025)

# wem_esoo_demand <- read_excel(wem_esoo_2024_operational_file) %>% 
#     clean_names() %>% 
#     #here, we select the observed and actual  
#     filter(scenario %in% c('Actual', 'Expected (Step Change)'),
#            parent_category == 'Operational (Sent Out)',
#            category != "Operational (Sent Out)",
#            region == 'SWIS') %>% 
#   group_by(year) %>% 
#   summarise(annual_consumption_t_wh = sum(annual_consumption_t_wh)) %>% 
#   filter(year >= 2025)

```

Load progressive change emissions intensity:

```{r}
#| echo: false
#| output: false
prog_change_ems <- read_excel('Data/2024 ISP generation and storage outlook/Core/2024 ISP - Progressive Change - Core.xlsx', 
                           sheet = 'Summary',
                           range = "H210:AK211") %>% 
  clean_names() %>% 
  pivot_longer(cols = contains('20'), 
               names_to = 'year',
               values_to = 'mt_co2e') %>% 
  rename(scenario = 1) %>% 
  mutate(scenario = '2024 ISP - Progressive Change',
         year = as.numeric(sub(".*_(\\d{2})", "20\\1", year))) %>% 
  select(- total)


prog_change_gen <- read_excel('Data/2024 ISP generation and storage outlook/Core/2024 ISP - Progressive Change - Core.xlsx', 
                           sheet = 'Summary',
                           range = "H57:AJ72") %>% 
  clean_names() %>% 
  pivot_longer(cols = contains('20'), 
               names_to = 'year',
               values_to = 'twh_gen') %>% 
  rename(technology = 1) %>% 
  mutate(year = as.numeric(sub(".*_(\\d{2})", "20\\1", year)),
         type = case_when(str_detect(technology, 'coal')|
                            str_detect(technology, 'gas') ~ 'non-renewable',
                          str_detect(technology, '[w|W]ind')|
                            str_detect(technology, 'solar')|
                            str_detect(technology, 'PV')|
                            str_detect(technology, 'renewable')|
                            str_detect(technology, 'Hydro')|
                            str_detect(technology, 'DSP') ~ 'renewable',
                          .default = 'storage'))

prog_change_intensity <- prog_change_gen %>% 
  filter(type != "storage") %>% 
  group_by(year) %>% 
  summarise(twh_gen = sum(twh_gen)) %>% 
  full_join(prog_change_ems) %>% 
  mutate(mt_co2e_twh = mt_co2e / twh_gen) %>% 
  select(year, scenario, mt_co2e_twh)
  
```

Load emissions intensities for CCA:

```{r}
#| echo: false

cca_path <- "Data/20240911_EP2024-4366_Modelling_Sectoral_Pathways_to_Net_Zero_Emissions_Section 2_Sectors_Charts (1).xlsx"

cca_intensities <- read_excel(cca_path, 
                              sheet = "Fig 13",
                              skip = 6) %>% 
  rename(scenario = 1) %>% 
  pivot_longer(cols = contains("20"), names_to = "year", values_to = "mt_co2e_twh") %>% 
  mutate(year = as.numeric(year))


#interpolate data

cca_intensities_interpolated <- cca_intensities %>%
    complete(scenario, year = seq(2025, 2050, by = 1)) %>%
    # Group by scenario to interpolate within each scenario
    group_by(scenario) %>%
    # Use approx to linearly interpolate missing values
    arrange(year) %>%
    mutate(mt_co2e_twh = approx(x = year[!is.na(mt_co2e_twh)], 
                                y = mt_co2e_twh[!is.na(mt_co2e_twh)], 
                                xout = year)$y) %>%
    ungroup()


```

\
\
Calculate emissions budget:

```{r}
#| echo: false

nem_emissions_cca <- full_join(cca_intensities_interpolated, esoo_demand, by = join_by(year)) %>% 
  mutate(mt_co2e = annual_consumption_t_wh * mt_co2e_twh)


emissions_budget_cca <- nem_emissions_cca %>% 
  filter(!is.na(scenario)) %>% 
  group_by(scenario) %>% 
  summarise(budget = sum(mt_co2e))

emissions_budget_cca

emissions_budget_cca_A40_G1.5 <- emissions_budget_cca %>% 
  filter(scenario == "A40/G1.5") %>% 
  pull(budget) 

```

Calculate emissions in hypothetical scenario where we get 2024 ESOO central demand and ISP Progressive Change emissions intensity:

```{r}
hypothetical_2030 <- esoo_demand %>% 
  full_join(prog_change_intensity, by = join_by(year)) %>% 
  mutate(mt_co2e = annual_consumption_t_wh * mt_co2e_twh,
         scenario = "Progressive change to 2030")

```

Create a post-2030 trajectory for our hypothetical scenario which matches the overall CCA A40/G1.5 2025-50 budget exactly and the trajectory as closely as possible.

```{r}

create_post_2030_traj <- function(emissions_budget, scenario_emissions, nem_emissions_cca){
  
  remaining_budget <- emissions_budget - scenario_emissions %>%
    filter(year <= 2029) %>%
    pull(mt_co2e) %>%
    sum()
  
  #create a reasonable value for 2030 based on 2025-29 decline rate
  current_decline_rate <- (1 - scenario_emissions %>% 
  filter(year == 2029) %>% 
  pull(mt_co2e) /
  scenario_emissions %>% 
  filter(year == 2025) %>% 
  pull(mt_co2e)) / 4
  
  
  ems_2030 <- scenario_emissions %>%  
    filter(year == 2030) %>% 
    pull(mt_co2e) 
  
  ems_2050 <- nem_emissions_cca %>%  
    filter(year == 2050,
           scenario == "A40/G1.5") %>% 
    pull(mt_co2e) 
  
  
  area_under_curve <- function(ems_2030, ems_2050, x) {
  
   area_2030_x <- (x - 2029) * (ems_2030 + ems_2050)/2
  
   area_x_2050 <- (2050 - x) * (ems_2050)
  
   area <- area_2030_x + area_x_2050
  
   return(area)
 }

 # Create a function that returns the difference between area and C
 diff_from_target <- function(x){abs(area_under_curve(ems_2030, ems_2050, x) - remaining_budget)}


 min_year <- optimize(f = diff_from_target, interval = c(2031, 2050))

 
 trajectory <- tibble(year = c(2030, round(min_year$minimum), 2050),
                      mt_co2e = c(ems_2030, ems_2050, ems_2050)) %>% 
   complete(year = seq(2025, 2050, by = 1)) %>%
   #interpolate
   arrange(year) %>%
   mutate(mt_co2e = approx(x = year[!is.na(mt_co2e)], 
                                y = mt_co2e[!is.na(mt_co2e)], 
                                xout = year)$y,
          scenario = "Emissions constraint trajectory")
  
 return(list(trajectory = trajectory, min_year = min_year))
}


result <- create_post_2030_traj(emissions_budget_cca_A40_G1.5, hypothetical_2030, nem_emissions_cca)

post_2030_trajectory <-  result$trajectory

min_year <- result$min_year$minimum

```

```{r}

joined_data <- nem_emissions_cca %>% 
  bind_rows(hypothetical_2030 %>% 
              filter(year <= 2030), 
            post_2030_trajectory) %>% 
  filter(scenario != "A50/G2") %>% 
  select(- c(mt_co2e_twh, annual_consumption_t_wh)) %>% 
  left_join(esoo_demand) %>% 
  mutate(mt_co2e_twh = mt_co2e / annual_consumption_t_wh)

emissions_plot <- joined_data %>% 
  ggplot(aes(x = year, y = mt_co2e, fill = scenario, colour = scenario)) +
  geom_line(size = 1) +
  theme_grattan() +
  geom_richlegend(aes(label = scenario)) +
  labs(title = "Emissions over time",
       subtitle = "Mt co2-e",
       x = "")

grattan_save_pptx("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/hypothetical_emissions_plot_progressive.pptx",
            p = emissions_plot,
            type = "normal")

intensity_plot <- joined_data %>% 
  ggplot(aes(x = year, y = mt_co2e_twh, fill = scenario, colour = scenario)) +
  geom_line(size = 1) +
  theme_grattan() +
  geom_richlegend(aes(label = scenario)) +
  labs(title = "Emissions intensity over time",
       subtitle = "Mt co2-e per TWh",
       x = "")

# grattan_save_pptx("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/hypothetical_intensity_plot_progressive.pptx",
#             p = intensity_plot,
#             type = "normal")

# grattan_save_all("Output/Atlas/hypothetical_intensity_plot_progressive.pdf",
#              object = intensity_plot)


joined_data %>% 
  filter(scenario == "A40/G1.5") %>% 
  pull(mt_co2e) %>% 
  sum()

joined_data %>% 
  filter(scenario != "A40/G1.5",
         !(scenario == 'Progressive change to 2030' & year == 2030),
         !is.na(mt_co2e)) %>% 
  pull(mt_co2e) %>% 
  sum()

```

**No QC needed below here\
**

Alternative way where we do straightline decline from 2029 (ignorance of 2030 value):

```{r}

create_post_2030_traj <- function(emissions_budget, scenario_emissions, nem_emissions_cca){
  
  remaining_budget <- emissions_budget - scenario_emissions %>%
    filter(year <= 2029) %>%
    pull(mt_co2e) %>%
    sum()
  
  # #create a reasonable value for 2030 based on 2025-29 decline rate
  # current_decline_rate <- (1 - scenario_emissions %>% 
  # filter(year == 2029) %>% 
  # pull(mt_co2e) /
  # scenario_emissions %>% 
  # filter(year == 2025) %>% 
  # pull(mt_co2e)) / 4
  # 
  
  ems_2029 <- scenario_emissions %>%  
    filter(year == 2029) %>% 
    pull(mt_co2e) 
  
  ems_2050 <- nem_emissions_cca %>%  
    filter(year == 2050,
           scenario == "A40/G1.5") %>% 
    pull(mt_co2e) 
  
  
  area_under_curve <- function(ems_2029, ems_2050, x) {
  
   area_2030_x <- (x - 2028) * (ems_2029 + ems_2050)/2
  
   area_x_2050 <- (2050 - x) * (ems_2050)
  
   area <- area_2030_x + area_x_2050 - ems_2029
  
   return(area)
 }

 # Create a function that returns the difference between area and C
 diff_from_target <- function(x){abs(area_under_curve(ems_2029, ems_2050, x) - remaining_budget)}


 min_year <- optimize(f = diff_from_target, interval = c(2031, 2050))

 straight_line <- function(x, min_year){
   
   m <- (ems_2050 - ems_2029) / (min_year$minimum - 2029)
   
   return(ems_2029 + m*(x - 2029))
 }
 
 
 trajectory <- tibble(year = c(2030, round(min_year$minimum), 2050),
                      mt_co2e = c(straight_line(2030, min_year), ems_2050, ems_2050)) %>% 
   complete(year = seq(2025, 2050, by = 1)) %>%
   #interpolate
   arrange(year) %>%
   mutate(mt_co2e = approx(x = year[!is.na(mt_co2e)], 
                                y = mt_co2e[!is.na(mt_co2e)], 
                                xout = year)$y,
          scenario = "Emissions constraint trajectory")
  
 return(list(trajectory = trajectory, min_year = min_year))
}


result <- create_post_2030_traj(emissions_budget_cca_A40_G1.5, hypothetical_2030, nem_emissions_cca)

post_2030_trajectory <-  result$trajectory

min_year <- result$min_year$minimum


```

Load emissions intensities according to CW government projections:

```{r}
#| echo: false

cw_projections_file <- "Data/australias-emissions-projections-2024-chart-data.xlsx"

bsline_electricity_ems <- read_excel(cw_projections_file,
                                shee = 'Figure 5',
                                range = "A3:AZ12") %>% 
  clean_names() %>% 
  rename(sector = 1) %>% 
  pivot_longer(cols = matches('\\d\\d\\d\\d'),
               names_to = 'year',
               values_to = 'emissions') %>% 
  mutate(year = parse_number(year)) %>% 
  filter(sector == 'Electricity') 

bsline_electricity_demand <- read_excel(cw_projections_file,
                                shee = 'Figure 19',
                                range = "A3:V12") %>% 
  clean_names() %>% 
  rename(fuel = 1) %>% 
  #remove storage
  filter(!str_detect(fuel, 'storage')) %>% 
  pivot_longer(cols = matches('\\d\\d\\d\\d'),
               names_to = 'year',
               values_to = 'annual_consumption_t_wh') %>% 
  mutate(year = parse_number(year)) %>% 
  group_by(year) %>% 
  summarise(annual_consumption_t_wh = sum(annual_consumption_t_wh))

bsline_intensity <- full_join(bsline_electricity_ems, bsline_electricity_demand) %>% 
  mutate(mt_co2e_twh = emissions / annual_consumption_t_wh,
         scenario = "Projected") %>% 
  select(-c(sector, emissions, annual_consumption_t_wh))
```

Climate works intensities:

```{r}
climworks_intensities_file <- "Data/climateworks_trajectories_!!!rough_numbers_!!!.xlsx"

climworks_intensities <- read_excel(climworks_intensities_file,
                                sheet = 'Sheet1',
                                skip = 1) %>% 
  clean_names() %>% 
  mutate(mt_co2e_twh = mt_co2e / consumption_twh) %>% 
  select(year, scenario, mt_co2e_twh) 


```

```{r}
#| echo: false

national_intensities <- bind_rows(cca_intensities, bsline_intensity, climworks_intensities) %>% 
  filter(year >= 2020)

national_intensities_interpolated <- national_intensities %>%
    complete(scenario, year = seq(2025, 2050, by = 1)) %>%
    # Group by scenario to interpolate within each scenario
    group_by(scenario) %>%
    # Use approx to linearly interpolate missing values
    arrange(year) %>%
    mutate(mt_co2e_twh = approx(x = year[!is.na(mt_co2e_twh)], 
                                y = mt_co2e_twh[!is.na(mt_co2e_twh)], 
                                xout = year)$y) %>%
    ungroup()

national_intensities_interpolated %>% 
  mutate(scenario = factor(scenario, levels = c("CW 1.5", "CW < 2", "A40/G1.5", "A50/G2", "Projected")) %>% fct_rev()) %>% 
  ggplot(aes(x = year, y = mt_co2e_twh, fill = scenario, colour = scenario)) +
  geom_line() +
  theme_grattan() +
  geom_richlegend(aes(label = scenario)) +
  geom_linepoint() +
  labs(title = "Climateworks has a much steeper trajectory than projected emissions",
       subtitle = "Mt co2-e per TWh")



  #geom_finallabel(aes(label = scenario)) 

```

```{r}
nem_emissions <- full_join(national_intensities_interpolated, esoo_demand) %>% 
  mutate(emissions = annual_consumption_t_wh * mt_co2e_twh)


emissions_budget_25_30 <- nem_emissions %>% 
  filter(!is.na(scenario),
         year >= 2025,
         year<= 2030) %>% 
  group_by(scenario) %>% 
  summarise(budget_2025_2030 = sum(emissions))

emissions_budget_25_30
```
