---
title: "Chapter charts"
author: "Dominic  Jones"
format: html
editor: visual
---

```{r}
#| echo: false
#setup

#load packages
library(targets)
library(dplyr)
library(tidyr) 
library(ggplot2)
#library(waterfalls)

```

# Households are set to save in the transition: fixing the safeguard will cost them little

## 1.Retail prices will look like this for each of the options:

## 2. But electricity is just one part of the household energy bill  

Average consumer energy spending by source

```{r}
#| echo: false

#filter out WA for now bcos no feed in data
#INCLUDE LATER

#load data
tar_load(average_electricity_costs)
tar_load(average_gas_costs)
tar_load(average_petrol_costs)
tar_load(household_connections)

chart_palette <- c(
  "Petrol" = grattan_red,
  "Gas" = grattan_orange, 
  "Electricity" = grattan_yellow,
  "Total costs without electrification" = grattan_black
)


net_average_electricity_costs <- average_electricity_costs %>% 
  mutate(category = "Electricity") %>% 
  group_by(year, state, electrification, category) %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  ungroup()

bind_rows(net_average_electricity_costs, 
          average_gas_costs, 
          average_petrol_costs) %>% 
  left_join(household_connections, by = join_by(year, state)) %>% 
  group_by(year, electrification, category) %>% 
  filter(year == 2025,
         electrification == T,
         state != "WA") %>% 
  summarise(average_cost_dollars = weighted.mean(average_cost_dollars, 
                                                 connections)) %>% 
  ggplot(aes(reorder(x = category,average_cost_dollars), 
             y = average_cost_dollars, fill = category, colour = category)) +
  geom_col() +
  theme_grattan() +
  scale_colour_manual(values = chart_palette) +
  scale_fill_manual(values = chart_palette) +
  grattan_y_continuous(labels = scales::dollar_format()) +
    labs(title = "Electricity forms a small part of total household energy costs",
       subtitle = 'Average annual household energy costs, by fuel source, 2025',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")
  

```

Save chart:

```{r}
#| echo: false

grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/average_energy_wallet.pdf",
             object = ggplot2::last_plot())
```

## 3. Households will save on energy as they electrify

```{r}
#| echo: false
chart_data <- bind_rows(net_average_electricity_costs, average_gas_costs, average_petrol_costs) %>% 
  left_join(household_connections) %>% 
  group_by(year, electrification, category) %>% 
  filter(year >= 2025,
         year <= 2043,
         state != "WA") %>% 
  summarise(average_cost_dollars = weighted.mean(average_cost_dollars, connections)) %>% 
  mutate(category = fct(category, levels = c("Petrol", "Electricity", "Gas"))) %>% 
  ungroup()



label_data <- chart_data %>% 
  filter(electrification == T) %>% 
  # Calculate the y positions for the labels at the midpoints of each stack
  group_by(year) %>% 
  arrange(category) %>% 
  mutate(y_position = cumsum(average_cost_dollars),
         y_position = y_position - 0.5 * average_cost_dollars) %>% 
  ungroup() %>% 
  bind_rows(chart_data %>% 
  filter(electrification == F) %>% 
    group_by(year) %>% 
    summarise(y_position = sum(average_cost_dollars)) %>% 
    ungroup() %>% 
    mutate(category = "Total costs without electrification"))


gap_by_2040 <- chart_data %>% 
  filter(year == 2040) %>% 
  group_by(electrification) %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  pivot_wider(names_from = electrification, values_from = average_cost_dollars) %>%
  mutate(gap = `FALSE` - `TRUE`) %>% 
  pull(gap) %>% 
  signif(3)

plot <- chart_data %>% 
ggplot(aes(x = year, y = average_cost_dollars)) +
    geom_area(data = . %>% 
             filter(electrification == T), 
           aes(x = year, y = average_cost_dollars, 
               colour = fct_rev(category), fill = fct_rev(category))) +
    geom_line(data = . %>% 
             filter(electrification == F) %>% 
               group_by(year) %>% 
                summarise(average_cost_dollars = sum(average_cost_dollars)),
             aes(x = year, y = average_cost_dollars),
             colour = grattan_black,
             linetype = 2,
             linewidth = 2) +
    grattan_label(data = label_data %>% 
                    filter(year == 2043), 
                aes(x = year, 
                    y = y_position, 
                    label = str_wrap(category, 13),
                    colour = category),
                hjust = 0,
                nudge_x = 0.5) +
    grattan_y_continuous(labels = scales::dollar_format(), expand_top = 0.1) +
    scale_x_continuous_grattan(expand_right = 0.2,
                               breaks = seq(2025,2040, by = 5)) +
    scale_colour_manual(values = chart_palette) +
    scale_fill_manual(values = chart_palette) +
    theme_grattan() +
    labs(title = paste0("Electrification will save the average Australian household", " $", gap_by_2040, " a year by 2040"),
       subtitle = 'Average annual household energy costs, by fuel source',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")


```

save chart:

```{r}
#| echo: false

grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/average_bills_over_time.pdf",
             object = plot)
```

\
\
4. Unlike in 2012, households can now adapt

```{r}
#| echo: false

rm(list = ls())
source("R/helpers.R")

tar_load(cameo_electricity_costs)
tar_load(cameo_gas_costs)
tar_load(cameo_petrol_costs)
tar_load(household_connections)

cameo_cost_data <- bind_rows(cameo_electricity_costs, cameo_gas_costs) %>% 
  cross_join(tibble(ice = c(0,1,2))) %>% 
  bind_rows(cameo_petrol_costs)


electrification_data <- cameo_cost_data %>% 
  filter(year == 2025) %>%
  group_by(year, cooking, water_heating, space_heating, ev, pv, ice, state) %>% 
  summarise(total_cost_dollars = sum(annual_cost_dollars)) %>% 
  mutate(consumer_type = paste(cooking, water_heating, space_heating, ev, pv, ice, sep = "_")) %>% 
  filter(consumer_type %in% c("gas_gas_gas_0_FALSE_1", # all gas with an ICE vehicle
                              "gas_gas_gas_1_FALSE_0",
                              "gas_gas_gas_1_TRUE_0",
                              "electric_electric_electric_1_TRUE_0")) %>% 
  mutate(consumer_name = fct_case_when(consumer_type == "gas_gas_gas_0_FALSE_1" ~ "Fossil fuel home",
                                       consumer_type == "gas_gas_gas_1_FALSE_0" ~ "Switch to an EV",
                                       consumer_type == "gas_gas_gas_1_TRUE_0" ~ "Install solar",
                                       consumer_type == "electric_electric_electric_1_TRUE_0" ~ "Electrify gas") %>% fct_rev()) %>% 
  filter(state %nin% c("WA", "NT", "Aus")) %>% 
  ungroup()


# average_electrification_data <- electrification_data %>% 
#   left_join(household_connections) %>% 
#   filter(!is.na(connections)) %>% 
#   group_by(consumer_name) %>% 
#   summarise(total_cost_dollars = weighted.mean(total_cost_dollars, connections))


fall_chart_data <- electrification_data %>% 
  select(state, consumer_name, total_cost_dollars) %>% 
  group_by(state) %>% 
  arrange(consumer_name) %>% 
  mutate(chart_value = if_else(consumer_name %in% c("Fossil fuel home"), total_cost_dollars, total_cost_dollars - lead(total_cost_dollars))) %>% 
  filter(state == "NSW and ACT") %>% 
  ungroup() %>% 
  select(consumer_name, chart_value) %>% 
  arrange(fct_rev(consumer_name))
  
  
  # Get the category names for axis labels
axis_labels <- c("Fossil fuel home",  "Fully electrified home")
chart_label_data <- fall_chart_data %>% 
  mutate(y_value = cumsum(chart_value),
         y_value = y_value - chart_value /2,
         x_value = row_number())

saving_pct <- fall_chart_data %>% 
  mutate(pct = (chart_value - sum(chart_value))/chart_value * 100) %>% 
  filter(consumer_name == "Fossil fuel home") %>% 
  pull(pct) %>% 
  signif(1)

bar_text <- fall_chart_data %>% 
  mutate(chart_value = if_else(chart_value >=0, 
                               paste0('$',signif(chart_value, 3)),
                               paste0('-$',-signif(chart_value, 3)))) %>% 
           pull(chart_value)

total_text <- paste0("$", fall_chart_data %>% 
                       pull(chart_value) %>% 
                       sum() %>% 
                       signif(3))

fall_chart_data %>% 
  waterfall(calc_total= TRUE, 
            rect_text_labels = bar_text,
            total_rect_text = total_text,
            total_rect_border_color = NA,
            rect_text_size = 2,
            rect_border = NA,
            theme_text_family = "theme_grattan",
            fill_colours = c(grattan_red,
                             grattan_orange,
                             grattan_orange,
                             grattan_orange),
            total_rect_color = grattan_yellow,
            fill_by_sign = F,
            total_axis_text = "Fully electrified home") +
  theme_grattan() +
  grattan_label(data = chart_label_data %>% 
                  filter(consumer_name != "Fossil fuel home"),
                aes(x = x_value + 1, y = y_value,
                    label = consumer_name),
                hjust = 0,
                nudge_x = -0.6) +
  grattan_y_continuous() +
  scale_x_discrete(breaks = c("Electrify gas","Switch to an EV"), labels = axis_labels) +
  labs(title = paste0("In NSW, households that electrify their homes can save", "more than ", saving_pct, " per cent on their bills"),
       subtitle = 'Annual household energy costs, by consumer type',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")


```

Save chart:

```{r}
#| echo: false
grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/wallet_waterfall.pdf",
             object = ggplot2::last_plot())
```

Version with no cars...

```{r}
#| echo: false

rm(list = ls())

tar_load(cameo_electricity_costs)
tar_load(cameo_gas_costs)
tar_load(cameo_petrol_costs)
tar_load(household_connections)

cameo_cost_data <- bind_rows(cameo_electricity_costs, cameo_gas_costs) %>% 
  cross_join(tibble(ice = c(0,1,2))) %>% 
  bind_rows(cameo_petrol_costs)


electrification_data <- cameo_cost_data %>% 
  filter(year == 2025) %>%
  group_by(year, cooking, water_heating, space_heating, ev, pv, ice, state) %>% 
  summarise(total_cost_dollars = sum(annual_cost_dollars)) %>% 
  mutate(consumer_type = paste(cooking, water_heating, space_heating, ev, pv, ice, sep = "_")) %>% 
  filter(consumer_type %in% c("gas_gas_gas_0_FALSE_0", # all gas with an ICE vehicle
                              "gas_gas_gas_0_TRUE_0",
                              "electric_electric_electric_0_TRUE_0")) %>% 
  mutate(consumer_name = fct_case_when(consumer_type == "gas_gas_gas_0_FALSE_0" ~ "Fossil fuel home",
                                       consumer_type == "gas_gas_gas_0_TRUE_0" ~ "Install solar",
                                       consumer_type == "electric_electric_electric_0_TRUE_0" ~ "Electrify gas") %>% fct_rev()) %>% 
  filter(state %nin% c("WA", "NT", "Aus")) %>% 
  ungroup()


# average_electrification_data <- electrification_data %>% 
#   left_join(household_connections) %>% 
#   filter(!is.na(connections)) %>% 
#   group_by(consumer_name) %>% 
#   summarise(total_cost_dollars = weighted.mean(total_cost_dollars, connections))


fall_chart_data <- electrification_data %>% 
  select(state, consumer_name, total_cost_dollars) %>% 
  group_by(state) %>% 
  arrange(consumer_name) %>% 
  mutate(chart_value = if_else(consumer_name %in% c("Fossil fuel home"), total_cost_dollars, total_cost_dollars - lead(total_cost_dollars))) %>% 
  filter(state == "NSW and ACT") %>% 
  ungroup() %>% 
  select(consumer_name, chart_value) %>% 
  arrange(fct_rev(consumer_name))
  
  
# Get the category names for axis labels
axis_labels <- c("Fossil fuel home", "", "", "Solar electric home")
chart_label_data <- fall_chart_data %>% 
  mutate(y_value = cumsum(chart_value),
         y_value = y_value - chart_value /2,
         x_value = row_number())


bar_text <- fall_chart_data %>% 
  mutate(chart_value = if_else(chart_value >=0, 
                               paste0('$',signif(chart_value, 3)),
                               paste0('-$',-signif(chart_value, 3)))) %>% 
           pull(chart_value)

total_text <- paste0("$", fall_chart_data %>% 
                       pull(chart_value) %>% 
                       sum() %>% 
                       signif(3))

fall_chart_data %>% 
  waterfall(calc_total= TRUE, 
            rect_text_labels = bar_text,
            total_rect_text = total_text,
            rect_text_size = 2,
            theme_text_family = "theme_grattan",
            fill_colours = c(grattan_red,
                             grattan_orange,
                             grattan_orange),
            total_rect_color = grattan_yellow,
            fill_by_sign = F,
            total_axis_text = "Solar electric home") +
  theme_grattan() +
  grattan_label(data = chart_label_data %>% 
                  filter(consumer_name != "Fossil fuel home"),
                aes(x = x_value + 1, y = y_value,
                    label = consumer_name),
                hjust = 0,
                nudge_x = -0.6) +
  grattan_y_continuous() +
  scale_x_discrete(labels = axis_labels, breaks = c(1,4)) +
  labs(title = "Households can decrease their energy bills",
       subtitle = 'Average annual household energy costs, by fuel source',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")

 

```

Try switching the order of change:

```{r}
rm(list = ls())
source("R/helpers.R")

tar_load(cameo_electricity_costs)
tar_load(cameo_gas_costs)
tar_load(cameo_petrol_costs)
tar_load(household_connections)

cameo_cost_data <- bind_rows(cameo_electricity_costs, cameo_gas_costs) %>% 
  cross_join(tibble(ice = c(0,1,2))) %>% 
  bind_rows(cameo_petrol_costs)


electrification_data <- cameo_cost_data %>% 
  filter(year == 2025) %>%
  group_by(year, cooking, water_heating, space_heating, ev, pv, ice, state) %>% 
  summarise(total_cost_dollars = sum(annual_cost_dollars)) %>% 
  mutate(consumer_type = paste(cooking, water_heating, space_heating, ev, pv, ice, sep = "_")) %>% 
  filter(consumer_type %in% c("gas_gas_gas_0_FALSE_1", # all gas with an ICE vehicle
                              "electric_electric_electric_0_FALSE_1",
                              "electric_electric_electric_0_TRUE_1",
                              "electric_electric_electric_1_TRUE_0")) %>% 
  mutate(consumer_name = fct_case_when(consumer_type == "gas_gas_gas_0_FALSE_1" ~ "Fossil fuel home",
                                       consumer_type == "electric_electric_electric_0_FALSE_1" ~ "Electrify gas",
                                       consumer_type == "electric_electric_electric_0_TRUE_1" ~ "Install solar",
                                       consumer_type == "electric_electric_electric_1_TRUE_0" ~ "Switch to an EV") %>% fct_rev()) %>% 
  filter(state %nin% c("WA", "NT", "Aus")) %>% 
  ungroup()


# average_electrification_data <- electrification_data %>% 
#   left_join(household_connections) %>% 
#   filter(!is.na(connections)) %>% 
#   group_by(consumer_name) %>% 
#   summarise(total_cost_dollars = weighted.mean(total_cost_dollars, connections))


fall_chart_data <- electrification_data %>% 
  select(state, consumer_name, total_cost_dollars) %>% 
  group_by(state) %>% 
  arrange(consumer_name) %>% 
  mutate(chart_value = if_else(consumer_name %in% c("Fossil fuel home"), total_cost_dollars, total_cost_dollars - lead(total_cost_dollars))) %>% 
  filter(state == "NSW and ACT") %>% 
  ungroup() %>% 
  select(consumer_name, chart_value) %>% 
  arrange(fct_rev(consumer_name))
  
  
  # Get the category names for axis labels
axis_labels <- c("Fossil fuel home",  "Fully electrified home")
chart_label_data <- fall_chart_data %>% 
  mutate(y_value = cumsum(chart_value),
         y_value = y_value - chart_value /2,
         x_value = row_number())

saving_pct <- fall_chart_data %>% 
  mutate(pct = (chart_value - sum(chart_value))/chart_value * 100) %>% 
  filter(consumer_name == "Fossil fuel home") %>% 
  pull(pct) %>% 
  signif(1)

bar_text <- fall_chart_data %>% 
  mutate(chart_value = if_else(chart_value >=0, 
                               paste0('$',signif(chart_value, 3)),
                               paste0('-$',-signif(chart_value, 3)))) %>% 
           pull(chart_value)

total_text <- paste0("$", fall_chart_data %>% 
                       pull(chart_value) %>% 
                       sum() %>% 
                       signif(3))

fall_chart_data %>% 
  waterfall(calc_total= TRUE, 
            rect_text_labels = bar_text,
            total_rect_text = total_text,
            total_rect_border_color = NA,
            rect_text_size = 1.5,
            rect_border = NA,
            theme_text_family = "theme_grattan",
            fill_colours = c(grattan_red,
                             grattan_orange,
                             grattan_orange,
                             grattan_orange),
            total_rect_color = grattan_yellow,
            fill_by_sign = F,
            total_axis_text = "Fully electrified home") +
  theme_grattan() +
  grattan_label(data = chart_label_data %>% 
                  filter(consumer_name != "Fossil fuel home"),
                aes(x = x_value + 1, y = y_value,
                    label = consumer_name),
                hjust = 0,
                nudge_x = -0.6) +
  grattan_y_continuous() +
  scale_x_discrete(breaks = c("Electrify gas","Switch to an EV"), labels = axis_labels) +
  labs(title = paste0("In NSW, households that electrify their homes can save", "more than ", saving_pct, " per cent on their bills"),
       subtitle = 'Annual household energy costs, by consumer type',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")


grattan_save_pptx("test_reverse_elec.pptx", ggplot2::last_plot())
```
