---
title: "Chapter charts"
author: "Dominic  Jones"
format: html
editor: visual
---

```{r}
#| echo: false
#setup

#load packages
library(targets)
library(dplyr)
library(tidyr) 
library(ggplot2)
library(patchwork)
library(grattantheme)
library(forcats)

#library(waterfalls)

```

# Households are set to save in the transition: fixing the safeguard will cost them little

## 1.Retail prices will look like this for each of the options:

## 2. But electricity is just one part of the household energy bill 

Average consumer energy spending by source

```{r}
#| echo: false

#filter out WA for now bcos no feed in data
#INCLUDE LATER

#load data
tar_load(weighted_average_electricity_costs)
tar_load(average_gas_costs)
tar_load(average_petrol_costs)
tar_load(household_connections)

chart_fill_palette <- c(
  "Petrol" = grattan_red,
  "Gas" = grattan_orange, 
  "Electricity" = grattan_yellow,
  "Electrification savings" = "transparent",
  "Total costs without electrification" = grattan_black
)

chart_colour_palette <- c(
  "Petrol" = grattan_red,
  "Gas" = grattan_orange, 
  "Electricity" = grattan_yellow,
  "Electrification savings" = grattan_black,
  "Total costs without electrification" = grattan_black
)


net_average_electricity_costs <- weighted_average_electricity_costs %>% 
  mutate(category = "Electricity") %>% 
  group_by(year, state, scenario, electrification, category) %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  ungroup()

bind_rows(net_average_electricity_costs, 
          average_gas_costs, 
          average_petrol_costs) %>% 
  left_join(household_connections, by = join_by(year, state)) %>% 
  group_by(year, electrification, category) %>% 
  filter(year == 2025,
         electrification == T,
         state != "WA") %>% 
  summarise(average_cost_dollars = weighted.mean(average_cost_dollars, 
                                                 connections)) %>% 
  ggplot(aes(reorder(x = category, -average_cost_dollars), 
             y = average_cost_dollars, fill = category, colour = category)) +
  geom_col() +
  theme_grattan() +
  scale_colour_manual(values = chart_fill_palette) +
  scale_fill_manual(values = chart_fill_palette) +
  grattan_y_continuous(labels = scales::dollar_format()) +
    labs(title = "Electricity forms a small part of total household energy costs",
       subtitle = 'Average annual household energy costs, by fuel source, 2025',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")
  

```

Save chart:

```{r}
#| echo: false

grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/average_energy_wallet.pdf",
             object = ggplot2::last_plot())
```

Electricity consumption state

```{r}
#| echo: false

tar_load(annual_electricity_consumption_averages)
tar_load(average_electricity_costs)

annual_electricity_consumption_averages %>% 
  mutate(consumer_type = paste0(pv, "_", battery)) %>% 
  left_join(average_electricity_costs %>% 
            select(state, year, consumer_type, prop)) %>% 
  filter(electrification == T,
         consumption_export == "Consumption") %>% 
  group_by(year, state) %>% 
  summarise(annual_consumption_kwh = weighted.mean(annual_consumption_kwh, prop))

tar_load(integrated_fuel_use)
tar_load(rbs_households)

integrated_fuel_use %>%
  filter(state == "Tas", 
         conversion == "unconverted") %>% 
  group_by(year, state, fuel) %>% 
  summarise(pj = sum(pj)) %>% 
  left_join(rbs_households) %>% 
  mutate(kwh = pj / occupied_households * 2.778e+8)
```

Electricity bill state

```{r}
#| echo: false

net_average_electricity_costs %>% 
  filter(electrification == T,
         year == 2025) %>%
  ggplot(aes(x = state , y = average_cost_dollars)) +
  geom_col()


```

dollar mwh state

```{r}
#| echo: false

average_electricity_costs %>% 
  filter(electrification == T,
         year == 2025) %>%
  group_by(year, state, consumer_type, prop) %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  left_join(annual_electricity_consumption_averages %>% 
              filter(consumption_export == "Consumption",
                     electrification == T) %>% 
              mutate(consumer_type = paste0(pv, "_", battery))) %>% 
  mutate(dollar_mwh = average_cost_dollars/ (annual_consumption_kwh / 1000)) %>% 
  group_by(year, state) %>% 
  summarise(dollar_mwh = weighted.mean(dollar_mwh, prop)) %>% 
  filter(state != "WA") %>% 
  ggplot(aes(x = state, y = dollar_mwh)) +
  geom_col(position = "dodge")


```

## 3. Households will save on energy as they electrify

```{r}
#| echo: false
chart_data <- bind_rows(net_average_electricity_costs, average_gas_costs, average_petrol_costs) %>% 
  left_join(household_connections) %>% 
  group_by(year, electrification, category) %>% 
  filter(year >= 2025,
         year <= 2043,
         state != "WA") %>% 
  summarise(average_cost_dollars = weighted.mean(average_cost_dollars, connections)) %>% 
  mutate(category = fct(category, levels = c("Petrol", "Electricity", "Gas"))) %>% 
  ungroup()



label_data <- chart_data %>% 
  filter(electrification == T) %>% 
  # Calculate the y positions for the labels at the midpoints of each stack
  group_by(year) %>% 
  arrange(category) %>% 
  mutate(y_position = cumsum(average_cost_dollars),
         y_position = y_position - 0.5 * average_cost_dollars) %>% 
  ungroup() %>% 
  bind_rows(chart_data %>% 
  filter(electrification == F) %>% 
    group_by(year) %>% 
    summarise(y_position = sum(average_cost_dollars)) %>% 
    ungroup() %>% 
    mutate(category = "Total costs without electrification"))


gap_by_2040 <- chart_data %>% 
  filter(year == 2040) %>% 
  group_by(electrification) %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  pivot_wider(names_from = electrification, values_from = average_cost_dollars) %>%
  mutate(gap = `FALSE` - `TRUE`) %>% 
  pull(gap) %>% 
  signif(3)

plot <- chart_data %>% 
ggplot(aes(x = year, y = average_cost_dollars)) +
    geom_area(data = . %>% 
             filter(electrification == T), 
           aes(x = year, y = average_cost_dollars, 
               colour = fct_rev(category), fill = fct_rev(category))) +
    geom_line(data = . %>% 
             filter(electrification == F) %>% 
               group_by(year) %>% 
                summarise(average_cost_dollars = sum(average_cost_dollars)),
             aes(x = year, y = average_cost_dollars),
             colour = grattan_black,
             linetype = 2,
             linewidth = 2) +
    grattan_label(data = label_data %>% 
                    filter(year == 2043), 
                aes(x = year, 
                    y = y_position, 
                    label = str_wrap(category, 13),
                    colour = category),
                hjust = 0,
                nudge_x = 0.5) +
    grattan_y_continuous(labels = scales::dollar_format(), expand_top = 0.1) +
    scale_x_continuous_grattan(expand_right = 0.2,
                               breaks = seq(2025,2040, by = 5)) +
    scale_colour_manual(values = chart_fill_palette) +
    scale_fill_manual(values = chart_fill_palette) +
    theme_grattan() +
    labs(title = paste0("On average, electrification will save Australian households", " $", gap_by_2040, " a year by 2040"),
       subtitle = 'Average annual household energy costs, by fuel source',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")

plot
```

save chart:

```{r}
#| echo: false

grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/average_bills_over_time.pdf",
             object = plot)
```

Average state bills over time, state

```{r}

line_chart_data <- bind_rows(net_average_electricity_costs, average_gas_costs, average_petrol_costs) %>% 
  group_by(year, state, electrification) %>% 
  filter(year >= 2025,
         year <= 2043,
         state != "WA") %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  ungroup()

line_chart_data %>% 
  ggplot(aes(x = year, y = average_cost_dollars, colour = state)) +
  geom_line(data = . %>% filter(electrification == T)) +
  # geom_line(data = . %>% filter(electrification == F),
  #           linetype = "dashed") +
  grattan_y_continuous(limits = c(2000, 8000),
                       labels = scales::dollar_format()) +
  grattan_x_continuous(expand_left = 0.1, expand_right = 0.1) +
  grattan_label(data = line_chart_data %>% 
                    filter(year == 2025,
                           electrification == T), 
                aes(x = year, 
                    y = average_cost_dollars, 
                    label = str_wrap(state, 13),
                    colour = state),
                hjust = 0,
                #nudge_y = position_jitter(),
                nudge_x = -0.5) +
  theme_grattan() +
  labs(title ="State average energy bills",
       subtitle = 'Average annual household energy costs, by fuel source',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")

```

```{r}
plot_data <- bind_rows(net_average_electricity_costs, average_gas_costs, average_petrol_costs) %>% 
  filter(year %in% c(2025,  2043),
         electrification == T) %>% 
  group_by(state, year) %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  filter(state != "WA") %>% 
  mutate(state = if_else(state == "NSW and ACT", "NSW", state))

state_levels <- plot_data %>% filter(year == 2025) %>% 
  arrange(average_cost_dollars) %>% 
  pull(state)

plot_data <-  plot_data %>% 
  mutate(state = fct(state, levels = state_levels))

arrows_single_data <- plot_data %>%
  group_by(state) %>%
  summarise(
    x_start = state[1],
    y_start = average_cost_dollars[year == 2025],
    x_end = state[1],
    y_end = average_cost_dollars[year == max(year)],
    .groups = "drop"
  )

plot_data %>% 
  ggplot(aes(reorder(x = state, average_cost_dollars), y = average_cost_dollars)) +
    geom_segment(data = arrows_single_data,
               aes(x = x_start, y = y_start, xend = x_end, yend = y_end + 30),
               arrow = grattan_arrow(),
               colour = grattan_grey1,
               size = 2) +
  grattan_point_filled(aes(color = year), size = 3, 
                       stroke = 2, fill = "white", shape = 21) +
    grattan_label(data= . %>% filter(state == "Qld"),
                aes(x = state,
                    y = average_cost_dollars,
                    label = year,
                    colour = year),
                hjust = 1,
                nudge_x = -0.2
                ) +
  grattan_y_continuous(limits = c(0,8000), labels = scales::dollar_format()) +
  scale_x_discrete(expand = expansion(add = c(0.9,0.5)))+
  theme_grattan() +
  labs(title ="Households across the country are set to save ",
       subtitle = 'Average annual household energy costs, by fuel source',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")
  


```

Bar chart version:

```{r}
#| echo: false
gap <- chart_data %>% 
  filter(year %in% c(2025, 2030, 2035, 2040)) %>% 
  group_by(year, electrification) %>% 
  summarise(average_cost_dollars = sum(average_cost_dollars)) %>% 
  pivot_wider(names_from = electrification, values_from = average_cost_dollars) %>%
  #create a row of costs for the counterfactual case
  mutate(average_cost_dollars = `FALSE` - `TRUE`,
         category = "Electrification savings",
         electrification = T) %>% 
  select(-c(`FALSE`, `TRUE`))


col_chart_data <- chart_data %>% 
  filter(year %in% c(2025, 2030, 2035, 2040),
         electrification == T) %>% 
  bind_rows(gap) %>% 
  mutate(category = fct(category, levels = c("Petrol", 
                                             "Electricity", 
                                             "Gas", 
                                             "Electrification savings"))) %>% 
  ungroup()


savings_box_data <- col_chart_data %>% 
  group_by(year) %>% 
  mutate(cumsum_top = cumsum(average_cost_dollars),
         cumsum_bottom = cumsum_top - average_cost_dollars) %>% 
  filter(category == "Electrification savings")


label_data <- col_chart_data %>% 
  # Calculate the y positions for the labels at the midpoints of each stack
  group_by(year) %>% 
  arrange(category) %>% 
  mutate(y_position = cumsum(average_cost_dollars),
         y_position = y_position - 0.5 * average_cost_dollars) %>% 
  ungroup()

plot <- col_chart_data %>% 
  filter(category != "Electrification savings") %>% 
ggplot(aes(x = year, y = average_cost_dollars)) +
    geom_col(aes(x = year, y = average_cost_dollars, 
               colour = fct_rev(category), fill = fct_rev(category))) +
    grattan_label(data = label_data %>% 
                    filter(year == 2040), 
                aes(x = year, 
                    y = y_position, 
                    label = str_wrap(category, 13),
                    colour = category),
                hjust = 0,
                nudge_x = 2.5) +
  
  
      # Add dotted borders for Electrification savings (top and sides only)
    geom_segment(data = savings_box_data,
                 aes(x = year - 2.25, xend = year + 2.25,
                     y = cumsum_top, yend = cumsum_top),  # top edge
                 linetype = "dashed", color = "black", size = 1) +
    geom_segment(data = savings_box_data,
                 aes(x = year - 2.25, xend = year - 2.25,
                     y = cumsum_bottom, yend = cumsum_top),  # left edge
                 linetype = "dashed", color = "black", size = 1) +
    geom_segment(data = savings_box_data,
                 aes(x = year + 2.25, xend = year + 2.25,
                     y = cumsum_bottom, yend = cumsum_top),  # right edge
                 linetype = "dashed", color = "black", size = 1) +
  
    grattan_y_continuous(labels = scales::dollar_format(), expand_top = 0.1) +
    scale_x_continuous_grattan(expand_right = 0.3,
                               breaks = seq(2025,2040, by = 5)) +
    scale_colour_manual(values = chart_colour_palette) +
    scale_fill_manual(values = chart_fill_palette) +
    theme_grattan() +
    labs(title = paste0("On average, electrification will save Australian households", " $", gap_by_2040, " a year by 2040"),
       subtitle = 'Average annual household energy costs, by fuel source',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")

plot
```

```{r}
grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/average_bills_time_columns.pdf",
             object = plot)

writeLines(as.character(gap_by_2040), "C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/average_electrification_savings_2040.txt")
```

\
\
4. Unlike in 2012, households can now adapt

```{r}
#| echo: false

rm(list = ls())
source("R/helpers.R")

tar_load(cameo_electricity_costs)
tar_load(cameo_gas_costs)
tar_load(cameo_petrol_costs)
tar_load(household_connections)

cameo_cost_data <- bind_rows(cameo_electricity_costs, cameo_gas_costs) %>% 
  cross_join(tibble(ice = c(0,1,2))) %>% 
  bind_rows(cameo_petrol_costs)


electrification_data <- cameo_cost_data %>% 
  filter(year == 2025) %>%
  group_by(year, cooking, water_heating, space_heating, ev, pv, ice, battery, state) %>% 
  summarise(total_cost_dollars = sum(annual_cost_dollars)) %>% 
  mutate(consumer_type = paste(cooking, water_heating, space_heating, ev, pv, battery, ice, sep = "_")) %>% 
  filter(consumer_type %in% c("gas_gas_gas_0_FALSE_FALSE_1", # all gas with an ICE vehicle
                              "gas_gas_gas_1_FALSE_FALSE_0",
                              "gas_gas_gas_1_TRUE_FALSE_0",
                              "gas_gas_gas_1_TRUE_TRUE_0",
                              "electric_electric_electric_1_TRUE_TRUE_0")) %>% 
  mutate(consumer_name = fct_case_when(consumer_type == "gas_gas_gas_0_FALSE_FALSE_1" ~ "Fossil fuel home",
                                       consumer_type == "gas_gas_gas_1_FALSE_FALSE_0" ~ "Switch to an EV",
                                       consumer_type == "gas_gas_gas_1_TRUE_FALSE_0" ~ "Install solar",
                                       consumer_type == "gas_gas_gas_1_TRUE_TRUE_0" ~ "Install a battery",
                                       consumer_type == "electric_electric_electric_1_TRUE_TRUE_0" ~ "Electrify gas") %>% fct_rev()) %>% 
  filter(state %nin% c("WA", "NT", "Aus")) %>% 
  ungroup()


# average_electrification_data <- electrification_data %>% 
#   left_join(household_connections) %>% 
#   filter(!is.na(connections)) %>% 
#   group_by(consumer_name) %>% 
#   summarise(total_cost_dollars = weighted.mean(total_cost_dollars, connections))


fall_chart_data <- electrification_data %>% 
  select(state, consumer_name, total_cost_dollars) %>% 
  group_by(state) %>% 
  arrange(consumer_name) %>% 
  mutate(chart_value = if_else(consumer_name %in% c("Fossil fuel home"), total_cost_dollars, total_cost_dollars - lead(total_cost_dollars))) %>% 
  filter(state == "NSW and ACT") %>% 
  ungroup() %>% 
  select(consumer_name, chart_value) %>% 
  arrange(fct_rev(consumer_name))
  
  
  # Get the category names for axis labels
axis_labels <- c("Fossil fuel home",  "Fully electrified home")
chart_label_data <- fall_chart_data %>% 
  mutate(y_value = cumsum(chart_value),
         y_value = y_value - chart_value /2,
         x_value = row_number())

saving_pct <- fall_chart_data %>% 
  mutate(pct = (chart_value - sum(chart_value))/chart_value * 100) %>% 
  filter(consumer_name == "Fossil fuel home") %>% 
  pull(pct) %>% 
  signif(1)

bar_text <- fall_chart_data %>% 
  mutate(chart_value = if_else(chart_value >=0, 
                               paste0('$',signif(chart_value, 3)),
                               paste0('-$',-signif(chart_value, 3)))) %>% 
           pull(chart_value)

total_text <- paste0("$", fall_chart_data %>% 
                       pull(chart_value) %>% 
                       sum() %>% 
                       signif(3))

fall_chart_data %>% 
  waterfall(calc_total= TRUE, 
            rect_text_labels = bar_text,
            total_rect_text = total_text,
            total_rect_border_color = NA,
            rect_text_size = 2,
            rect_border = NA,
            theme_text_family = "theme_grattan",
            fill_colours = c(grattan_red,
                             grattan_orange,
                             grattan_orange,
                             grattan_orange,
                             grattan_orange),
            total_rect_color = grattan_yellow,
            fill_by_sign = F,
            total_axis_text = "Fully electrified home") +
  theme_grattan() +
  grattan_label(data = chart_label_data %>% 
                  filter(consumer_name != "Fossil fuel home"),
                aes(x = x_value + 1, y = y_value,
                    label = consumer_name),
                hjust = 0,
                nudge_x = -0.6) +
  grattan_y_continuous(labels = scales::dollar_format()) +
  scale_x_discrete(breaks = c("Electrify gas","Switch to an EV"), labels = axis_labels) +
  labs(title = paste0("In NSW, households that electrify their homes can save", "more than ", saving_pct, " per cent on their bills"),
       subtitle = 'Annual household energy costs, by consumer type',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")


```

Save chart:

```{r}
#| echo: false
grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/wallet_waterfall.pdf",
             object = ggplot2::last_plot())
```

Attempt to facet:

```{r}
#| echo: false


electrification_data <- cameo_cost_data %>% 
  filter(year == 2025) %>%
  group_by(year, cooking, water_heating, space_heating, ev, pv, ice, state) %>% 
  summarise(total_cost_dollars = sum(annual_cost_dollars)) %>% 
  mutate(consumer_type = paste(cooking, water_heating, space_heating, ev, pv, ice, sep = "_")) %>% 
  filter(consumer_type %in% c("gas_gas_gas_0_FALSE_FALSE_1", # all gas with an ICE vehicle
                              "gas_gas_gas_1_FALSE_FALSE_0",
                              "gas_gas_gas_1_TRUE_FALSE_0",
                              "gas_gas_gas_1_TRUE_TRUE_0",
                              "electric_electric_electric_1_TRUE_0")) %>% 
  mutate(consumer_name = fct_case_when(consumer_type == "gas_gas_gas_0_FALSE_FALSE_1" ~ "Fossil fuel home",
                                       consumer_type == "gas_gas_gas_1_FALSE_0" ~ "Switch to an EV",
                                       consumer_type == "gas_gas_gas_1_TRUE_0" ~ "Install solar",
                                       consumer_type == "electric_electric_electric_1_TRUE_0" ~ "Electrify gas") %>% fct_rev()) %>% 
  filter(state %nin% c("WA", "NT", "Aus")) %>% 
  ungroup()


# average_electrification_data <- electrification_data %>% 
#   left_join(household_connections) %>% 
#   filter(!is.na(connections)) %>% 
#   group_by(consumer_name) %>% 
#   summarise(total_cost_dollars = weighted.mean(total_cost_dollars, connections))


fall_chart_data <- electrification_data %>% 
  select(state, consumer_name, total_cost_dollars) %>% 
  group_by(state) %>% 
  arrange(consumer_name) %>% 
  mutate(chart_value = if_else(consumer_name %in% c("Fossil fuel home"), total_cost_dollars, total_cost_dollars - lead(total_cost_dollars))) %>% 
  ungroup() %>% 
  select(state, consumer_name, chart_value) %>% 
  arrange(fct_rev(consumer_name))
  
  
  # Get the category names for axis labels
axis_labels <- c("Fossil fuel home",  "Fully electrified home")
chart_label_data <- fall_chart_data %>% 
  group_by(state) %>% 
  mutate(y_value = cumsum(chart_value),
         y_value = y_value - chart_value /2,
         x_value = row_number())

# saving_pct <- fall_chart_data %>% 
#   group_by(state) %>% 
#   mutate(pct = (chart_value - sum(chart_value))/chart_value * 100) %>% 
#   filter(consumer_name == "Fossil fuel home") %>% 
#   pull(pct) %>% 
#   signif(1)


create_plot <- function(data, label){
  
  
  bar_text <- data %>% 
  mutate(chart_value = if_else(chart_value >=0, 
                               paste0('$',signif(chart_value, 3)),
                               paste0('-$',-signif(chart_value, 3)))) %>% 
           pull(chart_value)

  total_text <- paste0("$", data %>% 
                       pull(chart_value) %>% 
                       sum() %>% 
                       signif(3))

data %>% 
    select(-state) %>% 
  waterfall(calc_total= TRUE, 
            rect_text_labels = bar_text,
            total_rect_text = total_text,
            total_rect_border_color = NA,
            rect_text_size = 2,
            rect_border = NA,
            theme_text_family = "theme_grattan",
            fill_colours = c(grattan_red,
                             grattan_orange,
                             grattan_orange,
                             grattan_orange),
            total_rect_color = grattan_yellow,
            fill_by_sign = F,
            total_axis_text = "Fully electrified home") +
  theme_grattan() +
  grattan_label(data = chart_label_data %>% 
                  filter(state == label,
                         consumer_name != "Fossil fuel home"),
                aes(x = x_value + 1, y = y_value,
                    label = consumer_name),
                hjust = 0,
                nudge_x = -0.6) +
  grattan_y_continuous(limits = c(0, 6200), labels = scales::dollar_format()) +
  scale_x_discrete(breaks = c("Electrify gas","Switch to an EV"), labels = axis_labels) +
ggtitle(label) +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold")) +
labs(x = "")
  
}


 # Split data by faceting variable
  split_data <- fall_chart_data %>% 
    #filter chart for no Tas for now
    filter(state!= "Tas") %>% 
    group_split(!!sym("state"))
  
  # Get the unique values for titles
  facet_labels <- fall_chart_data %>% 
    filter(state!= "Tas") %>% 
    distinct(!!sym("state")) %>% 
    pull(!!sym("state"))
  
  ncol <- 2
  
   plots <- map2(split_data, facet_labels, ~{
    base_plot <- create_plot( .x, .y)
     
     # Determine if this is a leftmost plot (first in each row)
    plot_index <- which(facet_labels == .y)
    is_leftmost <- (plot_index - 1) %% ncol == 0
    is_bottom <- plot_index > length(facet_labels) - ncol 
    
    if (!is_leftmost) {
      base_plot <- base_plot +
        theme(axis.text.y = element_blank(),
              axis.ticks.y = element_blank(),
              axis.text.x = element_blank(),
              axis.ticks.x = element_blank())
    }
    
    # if (!is_bottom) {
    #   base_plot <- base_plot +
    #     theme(axis.text.x = element_blank(),
    #           axis.ticks.x = element_blank())
    # }
    # 
    base_plot
     
  })
  
  # Combine plots
  wrap_plots(plots, ncol = ncol) 
    # plot_annotation(title = "Households in every state can access large savings from electrification",
    #              subtitle = "Annual household energy costs, by consumer type",
    #              caption = "Notes: \nSource: Grattan Institute analaysis see app X",
    #              theme = theme_grattan())
    
```

save chart:

```{r}
#| echo: false
grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/all_states_waterfall.pdf",
             object = ggplot2::last_plot())
```

Version with no cars...

```{r}
#| echo: false

electrification_data <- cameo_cost_data %>% 
  filter(year == 2025) %>%
  group_by(year, cooking, water_heating, space_heating, ev, pv, ice, state) %>% 
  summarise(total_cost_dollars = sum(annual_cost_dollars)) %>% 
  mutate(consumer_type = paste(cooking, water_heating, space_heating, ev, pv, ice, sep = "_")) %>% 
  filter(consumer_type %in% c("gas_gas_gas_0_FALSE_0", # all gas with an ICE vehicle
                              "gas_gas_gas_0_TRUE_0",
                              "electric_electric_electric_0_TRUE_0")) %>% 
  mutate(consumer_name = fct_case_when(consumer_type == "gas_gas_gas_0_FALSE_0" ~ "Fossil fuel home",
                                       consumer_type == "gas_gas_gas_0_TRUE_0" ~ "Install solar",
                                       consumer_type == "electric_electric_electric_0_TRUE_0" ~ "Electrify gas") %>% fct_rev()) %>% 
  filter(state %nin% c("WA", "NT", "Aus")) %>% 
  ungroup()


# average_electrification_data <- electrification_data %>% 
#   left_join(household_connections) %>% 
#   filter(!is.na(connections)) %>% 
#   group_by(consumer_name) %>% 
#   summarise(total_cost_dollars = weighted.mean(total_cost_dollars, connections))


fall_chart_data <- electrification_data %>% 
  select(state, consumer_name, total_cost_dollars) %>% 
  group_by(state) %>% 
  arrange(consumer_name) %>% 
  mutate(chart_value = if_else(consumer_name %in% c("Fossil fuel home"), total_cost_dollars, total_cost_dollars - lead(total_cost_dollars))) %>% 
  filter(state == "NSW and ACT") %>% 
  ungroup() %>% 
  select(consumer_name, chart_value) %>% 
  arrange(fct_rev(consumer_name))
  
  
# Get the category names for axis labels
axis_labels <- c("Fossil fuel home",  "Solar electric home")
chart_label_data <- fall_chart_data %>% 
  mutate(y_value = cumsum(chart_value),
         y_value = y_value - chart_value /2,
         x_value = row_number())


bar_text <- fall_chart_data %>% 
  mutate(chart_value = if_else(chart_value >=0, 
                               paste0('$',signif(chart_value, 3)),
                               paste0('-$',-signif(chart_value, 3)))) %>% 
           pull(chart_value)

total_text <- paste0("$", fall_chart_data %>% 
                       pull(chart_value) %>% 
                       sum() %>% 
                       signif(3))

fall_chart_data %>% 
  waterfall(calc_total= TRUE, 
            rect_text_labels = bar_text,
            total_rect_text = total_text,
            rect_text_size = 2,
            theme_text_family = "theme_grattan",
            fill_colours = c(grattan_red,
                             grattan_orange,
                             grattan_orange),
            total_rect_color = grattan_yellow,
            fill_by_sign = F,
            total_axis_text = "Solar electric home") +
  theme_grattan() +
  grattan_label(data = chart_label_data %>% 
                  filter(consumer_name != "Fossil fuel home"),
                aes(x = x_value + 1, y = y_value,
                    label = consumer_name),
                hjust = 0,
                nudge_x = -0.6) +
  grattan_y_continuous() +
  scale_x_discrete(labels = axis_labels, breaks = c("Electrify gas","Solar electric home")) +
  labs(title = "Households can decrease their energy bills",
       subtitle = 'Average annual household energy costs, by fuel source',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")


```

Try switching the order of change:

```{r}
#| echo: false

electrification_data <- cameo_cost_data %>% 
  filter(year == 2025) %>%
  group_by(year, cooking, water_heating, space_heating, ev, pv, ice, state) %>% 
  summarise(total_cost_dollars = sum(annual_cost_dollars)) %>% 
  mutate(consumer_type = paste(cooking, water_heating, space_heating, ev, pv, ice, sep = "_")) %>% 
  filter(consumer_type %in% c("gas_gas_gas_0_FALSE_1", # all gas with an ICE vehicle
                              "electric_electric_electric_0_FALSE_1",
                              "electric_electric_electric_0_TRUE_1",
                              "electric_electric_electric_1_TRUE_0")) %>% 
  mutate(consumer_name = fct_case_when(consumer_type == "gas_gas_gas_0_FALSE_1" ~ "Fossil fuel home",
                                       consumer_type == "electric_electric_electric_0_FALSE_1" ~ "Electrify gas",
                                       consumer_type == "electric_electric_electric_0_TRUE_1" ~ "Install solar",
                                       consumer_type == "electric_electric_electric_1_TRUE_0" ~ "Switch to an EV") %>% fct_rev()) %>% 
  filter(state %nin% c("WA", "NT", "Aus")) %>% 
  ungroup()


# average_electrification_data <- electrification_data %>% 
#   left_join(household_connections) %>% 
#   filter(!is.na(connections)) %>% 
#   group_by(consumer_name) %>% 
#   summarise(total_cost_dollars = weighted.mean(total_cost_dollars, connections))


fall_chart_data <- electrification_data %>% 
  select(state, consumer_name, total_cost_dollars) %>% 
  group_by(state) %>% 
  arrange(consumer_name) %>% 
  mutate(chart_value = if_else(consumer_name %in% c("Fossil fuel home"), total_cost_dollars, total_cost_dollars - lead(total_cost_dollars))) %>% 
  filter(state == "NSW and ACT") %>% 
  ungroup() %>% 
  select(consumer_name, chart_value) %>% 
  arrange(fct_rev(consumer_name))
  
  
  # Get the category names for axis labels
axis_labels <- c("Fossil fuel home",  "Fully electrified home")
chart_label_data <- fall_chart_data %>% 
  mutate(y_value = cumsum(chart_value),
         y_value = y_value - chart_value /2,
         x_value = row_number())

saving_pct <- fall_chart_data %>% 
  mutate(pct = (chart_value - sum(chart_value))/chart_value * 100) %>% 
  filter(consumer_name == "Fossil fuel home") %>% 
  pull(pct) %>% 
  signif(1)

bar_text <- fall_chart_data %>% 
  mutate(chart_value = if_else(chart_value >=0, 
                               paste0('$',signif(chart_value, 3)),
                               paste0('-$',-signif(chart_value, 3)))) %>% 
           pull(chart_value)

total_text <- paste0("$", fall_chart_data %>% 
                       pull(chart_value) %>% 
                       sum() %>% 
                       signif(3))

fall_chart_data %>% 
  waterfall(calc_total= TRUE, 
            rect_text_labels = bar_text,
            total_rect_text = total_text,
            total_rect_border_color = NA,
            rect_text_size = 1.5,
            rect_border = NA,
            theme_text_family = "theme_grattan",
            fill_colours = c(grattan_red,
                             grattan_orange,
                             grattan_orange,
                             grattan_orange),
            total_rect_color = grattan_yellow,
            fill_by_sign = F,
            total_axis_text = "Fully electrified home") +
  theme_grattan() +
  grattan_label(data = chart_label_data %>% 
                  filter(consumer_name != "Fossil fuel home"),
                aes(x = x_value + 1, y = y_value,
                    label = consumer_name),
                hjust = 0,
                nudge_x = -0.6) +
  grattan_y_continuous() +
  scale_x_discrete(breaks = c("Electrify gas","Switch to an EV"), labels = axis_labels) +
  labs(title = paste0("In NSW, households that electrify their homes can save", "more than ", saving_pct, " per cent on their bills"),
       subtitle = 'Annual household energy costs, by consumer type',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")


#grattan_save_pptx("test_reverse_elec.pptx", ggplot2::last_plot())
```

```{r}
#| echo: false 

electrification_data <- cameo_cost_data %>% 
  filter(year == 2025) %>%
  group_by(year, cooking, water_heating, space_heating, ev, pv, ice, state) %>% 
  summarise(total_cost_dollars = sum(annual_cost_dollars)) %>% 
  mutate(consumer_type = paste(cooking, water_heating, space_heating, ev, pv, ice, sep = "_")) %>% 
  filter(consumer_type %in% c("gas_gas_gas_0_FALSE_1", # all gas with an ICE vehicle
                              "electric_electric_electric_0_FALSE_1",
                              "electric_electric_electric_1_FALSE_0",
                              "electric_electric_electric_1_TRUE_0")) %>% 
  mutate(consumer_name = fct_case_when(consumer_type == "gas_gas_gas_0_FALSE_1" ~ "Fossil fuel home",
                                       consumer_type == "electric_electric_electric_0_FALSE_1" ~ "Electrify gas",
                                       consumer_type == "electric_electric_electric_1_FALSE_0" ~ "Switch to an EV",
                                       consumer_type == "electric_electric_electric_1_TRUE_0" ~ "Install solar") %>% fct_rev()) %>% 
  filter(state %nin% c("WA", "NT", "Aus")) %>% 
  ungroup()


# average_electrification_data <- electrification_data %>% 
#   left_join(household_connections) %>% 
#   filter(!is.na(connections)) %>% 
#   group_by(consumer_name) %>% 
#   summarise(total_cost_dollars = weighted.mean(total_cost_dollars, connections))


fall_chart_data <- electrification_data %>% 
  select(state, consumer_name, total_cost_dollars) %>% 
  group_by(state) %>% 
  arrange(consumer_name) %>% 
  mutate(chart_value = if_else(consumer_name %in% c("Fossil fuel home"), total_cost_dollars, total_cost_dollars - lead(total_cost_dollars))) %>% 
  filter(state == "NSW and ACT") %>% 
  ungroup() %>% 
  select(consumer_name, chart_value) %>% 
  arrange(fct_rev(consumer_name))
  
  
  # Get the category names for axis labels
axis_labels <- c("Fossil fuel home",  "Fully electrified home")
chart_label_data <- fall_chart_data %>% 
  mutate(y_value = cumsum(chart_value),
         y_value = y_value - chart_value /2,
         x_value = row_number())

saving_pct <- fall_chart_data %>% 
  mutate(pct = (chart_value - sum(chart_value))/chart_value * 100) %>% 
  filter(consumer_name == "Fossil fuel home") %>% 
  pull(pct) %>% 
  signif(1)

bar_text <- fall_chart_data %>% 
  mutate(chart_value = if_else(chart_value >=0, 
                               paste0('$',signif(chart_value, 3)),
                               paste0('-$',-signif(chart_value, 3)))) %>% 
           pull(chart_value)

total_text <- paste0("$", fall_chart_data %>% 
                       pull(chart_value) %>% 
                       sum() %>% 
                       signif(3))

fall_chart_data %>% 
  waterfall(calc_total= TRUE, 
            rect_text_labels = bar_text,
            total_rect_text = total_text,
            total_rect_border_color = NA,
            rect_text_size = 1.5,
            rect_border = NA,
            theme_text_family = "theme_grattan",
            fill_colours = c(grattan_red,
                             grattan_orange,
                             grattan_orange,
                             grattan_orange),
            total_rect_color = grattan_yellow,
            fill_by_sign = F,
            total_axis_text = "Fully electrified home") +
  theme_grattan() +
  grattan_label(data = chart_label_data %>% 
                  filter(consumer_name != "Fossil fuel home"),
                aes(x = x_value + 1, y = y_value,
                    label = consumer_name),
                hjust = 0,
                nudge_x = -0.6) +
  grattan_y_continuous() +
  scale_x_discrete(breaks = c("Electrify gas","Switch to an EV"), labels = axis_labels) +
  labs(title = paste0("In NSW, households that electrify their homes can save", "more than ", saving_pct, " per cent on their bills"),
       subtitle = 'Annual household energy costs, by consumer type',
       x = '',
       y = '',
       caption = "Notes: \nSource: Grattan Institute analaysis see app X")

```

Different states have different opportunities to save:

```{r}
#| echo: false

electrification_data <- cameo_cost_data %>% 
  filter(year == 2025) %>%
  group_by(year, cooking, water_heating, space_heating, ev, pv, ice, state) %>% 
  summarise(total_cost_dollars = sum(annual_cost_dollars)) %>% 
  mutate(consumer_type = paste(cooking, water_heating, space_heating, ev, pv, ice, sep = "_")) %>% 
  filter(consumer_type %in% c("gas_gas_gas_0_FALSE_1", # all gas with an ICE vehicle
                              "electric_electric_electric_0_FALSE_1",
                              "electric_electric_electric_0_TRUE_1",
                              "electric_electric_electric_1_TRUE_0")) %>% 
  mutate(consumer_name = fct_case_when(consumer_type == "gas_gas_gas_0_FALSE_1" ~ "Fossil fuel home",
                                       consumer_type == "electric_electric_electric_0_FALSE_1" ~ "Electrify gas",
                                       consumer_type == "electric_electric_electric_0_TRUE_1" ~ "Install solar",
                                       consumer_type == "electric_electric_electric_1_TRUE_0" ~ "Switch to an EV") %>% fct_rev()) %>% 
  filter(state %nin% c("WA", "NT", "Aus")) %>% 
  ungroup()

savings_state <- electrification_data %>% 
  select(state, consumer_name, total_cost_dollars) %>% 
  group_by(state) %>% 
  arrange(consumer_name) %>% 
  mutate(savings = if_else(consumer_name %in% c("Fossil fuel home"), 0, lead(total_cost_dollars) - total_cost_dollars)) %>% 
filter(savings != 0)


savings_state %>% 
  ggplot(aes(x = state, y = savings, fill = consumer_name)) +
  geom_col()
```

Consumer type costs over time

```{r}
cameo_cost_data <- bind_rows(cameo_electricity_costs, cameo_gas_costs) %>% 
  cross_join(tibble(ice = c(0,1,2))) %>% 
  bind_rows(cameo_petrol_costs)


electrification_data <- cameo_cost_data %>% 
  filter(year %in% c(2025, 2030, 2035, 2040)) %>% 
  mutate(consumer_type = paste(cooking, water_heating, space_heating, ev, pv, ice, sep = "_")) %>% 
  filter(consumer_type %in% c("gas_gas_gas_0_FALSE_0", # all gas
                              "electric_electric_electric_0_FALSE_0")) %>%  #all electric
  mutate(consumer_name = fct_case_when(consumer_type == "gas_gas_gas_0_FALSE_0" ~ "Gas cooking, water heating and space heating",
                                       consumer_type == "electric_electric_electric_0_FALSE_0" ~ "All electric household") %>% fct_rev(),
         category = case_when(category == "Electricity consumption" ~ "Electricity", 
                              category == "Gas volumetric" ~ "Gas consumption",
                              .default = category),
         category = fct(category, 
                        levels = c("Electricity", 
                                   "Gas consumption", 
                                   "Gas connection",
                                   "Petrol")) %>% fct_rev()) %>% 
  select(year, consumer_name, state, category, annual_cost_dollars) %>%
  filter(state %in% c("Vic")) %>% 
  ungroup()

chart_fill_palette <- c(
  "Petrol" = grattan_red,
  "Gas consumption" = grattan_orange, 
  "Gas connection" = grattan_darkorange,
  "Electricity" = grattan_yellow
)

electrification_data %>% 
  ggplot(aes(x = year, y = annual_cost_dollars, fill = category)) +
  geom_col() +
  facet_wrap(~consumer_name) +
  grattan_y_continuous(limits = c(0, 4000),
                       labels = scales::dollar_format()) +
  theme_grattan() +
  scale_fill_manual(values = chart_fill_palette) +
  labs(title = "As connection charges grow, households that are stuck on gas will pay far more than they need to",
       subtitle = "Annual gas and electricity costs by household type",
       x = "",
       y= "")
```

Save chart

```{r}
#| echo: false
grattan_save_all("C:/Users/domijones/Grattan Institute Dropbox/Dominic  Jones/Apps/Overleaf/energy-2025-carbon-pricing-for-electricity/atlas/gas_v_electric_vic.pdf",
             object = ggplot2::last_plot())
```
